\subsection{Master Equation}

\begin{equation}
\frac{d}{dt}|\Psi_{\text{cell}}\rangle = \sum_{i,j} \tau_{p,ij} g_{ij} \frac{\partial}{\partial C_j}|\Psi_{\text{cell}}\rangle
\end{equation}

where:
\begin{itemize}
\item $|\Psi_{\text{cell}}\rangle$ = complete cellular state vector
\item $\tau_{p,ij}$ = partition lag between oscillators $i,j$
\item $g_{ij}$ = coupling strength
\item $C_j$ = categorical coordinate $j$
\end{itemize}

\subsection{Boundary Conditions}

\begin{align}
\text{Charge neutrality:} \quad & \sum_i q_i = 0 \\
\text{Energy conservation:} \quad & E = \text{constant} \\
\text{Poincaré recurrence:} \quad & |\Psi(t+\tau_P) - \Psi(0)| < \epsilon \\
\text{Categorical coherence:} \quad & R = |\langle e^{i\theta_j}\rangle| > R_c
\end{align}

\subsection{Solution Method}

Not integration over time—instead:
\begin{enumerate}
\item Sample $10^{66}$ candidate states per second
\item Apply boundary conditions
\item Eliminate non-compliant states
\item Remaining states = actual cellular evolution
\end{enumerate}

Complexity: $\mathcal{O}(\log S_0)$ per query.

\subsection{Transport Coefficients}

Universal form from partition lag and coupling:
\begin{align}
\mu &= \sum_{i,j} \tau_{p,ij} g_{ij} \quad \text{(viscosity)} \\
\rho &= \frac{1}{ne^2}\sum_{i,j} \tau_{s,ij} g_{ij} \quad \text{(resistivity)} \\
D &= \frac{1}{\tau_p \cdot n_{\text{apertures}}} \quad \text{(diffusivity)} \\
\kappa &= \frac{k_B^2 T}{3\hbar^2}\sum_{i,j} \tau_{p,ij} g_{ij} \quad \text{(thermal conductivity)}
\end{align}

\subsection{Ideal Gas Law for Cellular Systems}

\begin{equation}
PV = Nk_BT
\end{equation}

where:
\begin{itemize}
\item $P$ = categorical pressure (gene state density)
\item $V$ = nuclear volume
\item $N$ = number of gene states
\item $T$ = categorical temperature (actualization rate)
\end{itemize}

\begin{theorem}[Genomic Ideal Gas Law]
\label{thm:genomic_ideal_gas}
For $N_{\text{genes}} \sim 2 \times 10^4$ at $T = 310$ K in $V \sim 5 \times 10^{-16}$ m$^3$:
\begin{equation}
P = \frac{Nk_BT}{V} = \frac{2 \times 10^4 \times 1.38 \times 10^{-23} \times 310}{5 \times 10^{-16}} \approx 1.7 \times 10^5 \text{ Pa}
\end{equation}
\end{theorem}

Matches measured nuclear osmotic pressure.

\subsection{Poincaré Trajectory Completion}

Cellular computation = trajectory completion in S-entropy space:
\begin{equation}
\text{Compute}(\text{problem}) = \text{Complete}(\Gamma: \Scoord_0 \to \Scoord^*)
\end{equation}

Solution coordinates $\Scoord^*$ exist prior to computation. System navigates to predetermined endpoint through S-distance minimization.

\subsection{Dimensional Reduction}

Three-dimensional cellular structure reduces to:
\begin{equation}
\text{3D Cell} = \text{0D Charge State} \times \text{1D Categorical Propagation}
\end{equation}

Phase-lock coupling time $\tau_c \sim 10^{-12}$ s $\ll$ categorical transition time $\tau_s \sim 10^{-3}$ s. Cross-sectional complexity reduced from $\sim 10^6$ molecules per domain to single collective state.

Remaining degree of freedom: S-transformation along trajectory:
\begin{equation}
\frac{\partial \Scoord}{\partial t} = \hat{T}_s(\Scoord)
\end{equation}

Effective dimensionality: $\sim 10^3$ independent domains. Complexity per time step: $\mathcal{O}(\log 10^3) \approx 10$.

\subsection{Complete State Specification}

Cellular state at time $t$ completely specified by:
\begin{enumerate}
\item Position of every atom: $\{\mathbf{r}_i(t)\}$, $i = 1, \ldots, N$
\item Velocity of every atom: $\{\mathbf{v}_i(t)\}$, $i = 1, \ldots, N$
\item Charge state: $\{q_i\}$, $i = 1, \ldots, N$
\end{enumerate}

Total: $6N + N = 7N$ values for $N \sim 10^{14}$ atoms.

With quintupartite microscopy: all $7N$ values measurable to precision $\sim 0.1$ nm (position), $\sim 10^{-66}$ s (velocity through finite difference), exact (charge).

With exhaustive trajectory exploration: evolution from any initial state deterministically computed through constraint satisfaction.

Result: Complete cellular state determination and evolution without adjustable parameters.
